\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-11 tex/paper.tex
more on AVO and AVI-MC

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation} 
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the marginal distribution parameterized by $\bfpsi$
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}
Effectively, we are optimizing a variational posterior $q(\bftheta | \psi)$ by minimizing the 
divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
Consequently, the proposed algorithm is a form of variational Bayesian inference through model criticism~\citep{box1980sampling}.

@@

If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values. Very large penalties may 
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}

@@

\textbf{KC: could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}

>>>>

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \tilde{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \tilde{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation} 
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
%
%Obviously, the variational relaxation could similarly be coupled with
%other variants of GANs and/or of evolution strategies.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

\subsection{Adversarial variational inference through model criticism}

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the marginal distribution parameterized by $\bfpsi$
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in \citep{2016arXiv161009033R}, though more complicated than a simple reparameterization of normally distributed noise $\bfz$ through a differentiable function $g(\bfz, \bfpsi)$. In our case, the variational program is a marginalized, non-differentiable  simulator.  Its density is  intractable; nevertheless, it can generate samples for $\bfx$ and is differentiable with respect to $\bfpsi$ and $\bfphi$. 
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown joint density $p(\bfx,\bftheta)$.

Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.

@@

\subsection{Adversarial variational learning}

If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values. Very large penalties may 
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}

%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here analogous to their use in black box variational inference~\citep{ranganath2014black}.

\kcnote{Discuss hierarchical models and iid data x. Factor of $N/M$ for AVI-MC?}

\newpage
% ==============================================================================

@@

\kcnote{could also use this to tune non-differentialble fast simulator to slower full simulator. Then fixed data size not an issue}
